Fisher’s Method

### 2018-10-15 Fishers_method/Fishers_method_101518.Rmd
updated asymptotic results in appendix

Under the Neyman and Pearson hypothesis testing framework, the significance level follows a uniform distribution on $\lbrack0,1\rbrack$ when the null hypothesis holds, however, the exact distribution under the alternative hypothesis is unknown. As a consequence of this, we rely on long known results from asymptotic theory. Lambert and Hall have shown that, given the test statistic is asymptotically normal, the one sided P-value follows a lognormal distribution with mean $-nc(\theta)$ and variance $n\tau^2(\theta)$, where $n$ is the within group sample size (1982). The parameter $c(\theta)$ is defined as half the Bahadur slope, given by $-\frac{1}{n}lim_{n\to\infty}logP_n=c(\theta)$, and is the exponential rate at which the significance level converges to zero under the alternative hypothesis. In addition, observe that the variance of the standardized P-value is $\frac {\tau^2(\theta)}{2n}$. We can approximate the two sided P-value by doubling the one sided one, which implies that $P_n\sim AlogN\lbrack-\frac{1}{2}nc(\theta), \frac {1}{4}n\tau^2(\theta)\rbrack$, and thus $logP_n\sim AN\lbrack-\frac{1}{2} nc(\theta), \frac{1}{4}n\tau^2(\theta)\rbrack$. Multiplying $logP_n$ by $-2$, we obtain $-2logP_n\sim AN\lbrack nc(\theta), n\tau^2(\theta)\rbrack$. Note that $-2logP_{n_j}$, $j=1,...,k$, are independent random variables asymptotically following Normal distributions, therefore under $H_a$ $$X^2_n=-2\sum_{j=1}^{k} logP_{n_j} \sim AN(\sum_{j=1}^{k}n_j c_j(\theta), \sum_{j=1}^{k}n_j \tau^2_j(\theta))$$.

Now consider the two sample shift problem as an example. Let $Y_{11},Y_{21},...,Y_{n_11}$ denote a sample of i.i.d. observations from a normal distribution with mean $\mu$ and standard deviation 1. Let $Y_{12},Y_{22},...,Y_{n_22}$ denote a second sample, independent of the first, with i.i.d. observations from a normal $(\mu+\theta, 1)$. For simplicity, suppose that $n_1=n_2=n$. We test $H_0:\theta=0$ versus $H_a:\theta\neq0$ using the test statistic $\frac{\sqrt{n}(\overline{Y_2}-\overline{Y_1})}{\sqrt{2}}$. According to Lambert and Hall, $c(\theta)=\frac{1}{2}\lambda\overline{\lambda}\theta^2$ and $\tau^2(\theta)=\lambda\overline{\lambda}\theta^2$, where $\lambda$ denotes the fractional sample size, and $\overline{\lambda}=1-\lambda$. Since we are assuming $n_1=n_2$, both $\lambda$ and $\overline{\lambda}$ are $\frac{1}{2}$. Therefore, in the two-sample shift problem we have $$-2logP_{n} \sim AN(\frac{1}{4}n\theta^2, \frac{1}{2}n\theta^2).$$ Note that $\theta$ in this case can be interpreted as Cohen's d because $\sigma$ is assumed to be 1.

@@

The asymptotic results can confirm that the power of Fisher's method approaches 1 as the within study sample sizes go to infinity, which is consistent with the power simulations presented in the body of the paper. To see this, note that if all $k$ studies are testing the same hypothesis, it follows that $X^2_n$ is asymptotically normal with grand mean $\sum_{j=1}^{k}\frac{1}{2}n_j \lambda_j \overline{\lambda_j}\theta_j^2$ and grand variance $\sum_{j=1}^{k}n_j \lambda_j \overline{\lambda_j}\theta_j^2$. Consequently, the asymptotic power of Fisher's test is:

$$\begin{aligned}
Power&=Pr(reject \ H_0 \ | \ H_a \ is \ true)\\
&=1-Pr(fail \ to \ reject \ H_0 \ | \ H_a \ is \ true)\\
&=1-Pr(|X^2_n| \ < \ \chi^2_{2k} \ | \ H_a \ is \ true)\\
&=1-\Phi\Big(\frac{\chi^2_{2k}-\sum_{j=1}^{k} \frac{1}{2} n_j\lambda_j \overline{\lambda}_j\theta^2_j}{\sum_{j=1}^{k}n_j\lambda_j \overline{\lambda}_j\theta^2_j}\Big)+\Phi\Big(\frac{-\chi^2_{2k}-\sum_{j=1}^{k} \frac{1}{2} n_j\lambda_j \overline{\lambda}_j\theta^2_j}{\sum_{j=1}^{k}n_j\lambda_j \overline{\lambda}_j\theta^2_j}\Big)\\
&=1-\Phi\Big(\frac{\chi^2_{2k}}{\sum_{j=1}^{k}n_j\lambda_j \overline{\lambda}_j\theta^2_j}-\frac{1}{2}\Big)+\Phi\Big(\frac{-\chi^2_{2k}}{\sum_{j=1}^{k}n_j\lambda_j \overline{\lambda}_j\theta^2_j}-\frac{1}{2}\Big)\\
&\xrightarrow{}1-\Phi(-\frac{1}{2})+\Phi(-\frac{1}{2}) ~~ as ~~ n\rightarrow \infty\\
&=1
\end{aligned}.$$

>>>>

Under the Neyman and Pearson hypothesis testing framework, the significance level follows a uniform distribution on $\lbrack0,1\rbrack$ when the null hypothesis holds, however, the exact distribution under the alternative hypothesis is unknown. As a consequence of this, we rely on long known results from asymptotic theory. Lambert and Hall have shown that, given the test statistic is asymptotically normal, the one sided P-value follows a lognormal distribution with mean $-nc(\theta)$ and variance $n\tau^2(\theta)$, where $n$ is the within study sample size (1982). The parameter $c(\theta)$ is defined as half the Bahadur slope, given by $-\frac{1}{n}lim_{n\to\infty}logP_n=c(\theta)$, and is the exponential rate at which the significance level converges to zero under the alternative hypothesis. In addition,the variance of the standardized P-value is $\frac {\tau^2(\theta)}{2n}$.Now, given that $P_n$ is asymptotically lognormal, then $logP_n\sim AN(-nc(\theta), n\tau^2(\theta))$, and thus $-2logP_n\sim AN(2nc(\theta), 4n\tau^2(\theta))$. Note that $-2logP_{n_j}$, $j=1,...,k$, are asymptotically independent random variables, therefore $$X^2_n=-2\sum_{j=1}^{k} logP_{n_j} \sim AN(2\sum_{j=1}^{k}n_j c_j(\theta), 4\sum_{j=1}^{k}n_j \tau^2_j(\theta))$$ under $P_\theta$. 
Now consider the two sample shift problem as an example. Let $Y_{11},Y_{21},...,Y_{n_11}$ denote a sample of i.i.d. observations from a normal distribution with mean $\mu$ and standard deviation 1. Let $Y_{12},Y_{22},...,Y_{n_22}$ denote a second sample, independent of the first, with i.i.d. observations from a normal $(\mu+\theta, 1)$. For simplicity, suppose that $n_1=n_2=n$. We test $H_0:\theta=0$ versus $H_a:\theta\neq0$ using the test statistic $\frac{\sqrt{n}(\overline{Y_2}-\overline{Y_1})}{\sqrt{2}}$. According to Lambert and Hall, $c(\theta)=\frac{1}{2}\lambda\overline{\lambda}\theta^2$ and $\tau^2(\theta)=\lambda\overline{\lambda}\theta^2$, where $\lambda$ denotes the fractional sample size, and $\overline{\lambda}=1-\lambda$. Since we are assuming $n_1=n_2$, both $\lambda$ and $\overline{\lambda}$ are $\frac{1}{2}$. Therefore, in the two-sample shift problem we have $$-2logP_{n} \sim AN(\frac{1}{4}n\theta^2, n\theta^2).$$ Note that $\theta$ in this case can be interpreted as Cohen's d because $\sigma$ is assumed to be 1.

@@

The asymptotic results can confirm that the power of Fisher's method approaches 1 as the within study sample sizes go to infinity, which is consistent with the power simulations presented in the body of the paper. To see this, note that if all $k$ studies are testing the same hypothesis, it follows that $X^2_n$ is asymptotically normal with grand mean $\sum_{j=1}^{k} 4n_j \lambda_j \overline{\lambda_j}\theta_j^2$ and grand variance $\sum_{j=1}^{k}n_j \lambda_j \overline{\lambda_j}\theta_j^2$. Consequently, the asymptotic power of Fisher's test is:

$$\begin{aligned}
Power&=Pr(reject \ H_0 \ | \ H_a \ is \ true)\\
&=1-Pr(fail \ to \ reject \ H_0 \ | \ H_a \ is \ true)\\
&=1-Pr(|X^2_n| \ < \ \chi^2_{2k} \ | \ H_a \ is \ true)\\
&=1-\Phi\Big(\frac{\chi^2_{2k}-\sum_{j=1}^{k}  n_j\lambda_j \overline{\lambda}_j\theta^2_j}{\sqrt{\sum_{j=1}^{k}4n_j\lambda_j \overline{\lambda}_j\theta^2_j}} \Big)+\Phi\Big(\frac{-\chi^2_{2k}-\sum_{j=1}^{k}  n_j\lambda_j \overline{\lambda}_j\theta^2_j}{\sqrt{\sum_{j=1}^{k}4n_j\lambda_j \overline{\lambda}_j\theta^2_j}}\Big)\\
&=1-\Phi\Big(\frac{\chi^2_{2k}}{\sqrt{\sum_{j=1}^{k}4n_j\lambda_j \overline{\lambda}_j\theta^2_j}}-\frac{\sqrt{\sum_{j=1}^{k}n_j\lambda_j \overline{\lambda}_j\theta^2_j}}{2}\Big)+\Phi\Big(\frac{-\chi^2_{2k}}{\sqrt{\sum_{j=1}^{k}4n_j\lambda_j \overline{\lambda}_j\theta^2_j}}-\frac{\sqrt{\sum_{j=1}^{k}n_j\lambda_j \overline{\lambda}_j\theta^2_j}}{2}\Big)\\
&\xrightarrow{}1 ~~ as ~~ n\rightarrow \infty\\
\end{aligned}.$$
